---
title: "Statics Dashboards Using ggplot"
author: "Maya Reese Farmer"
date: "October 21, 2021"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
---

#### Introduction  
I'm going to be constructing 2 static dashboards using the Sales Conversion Optimization dataset from kaggle.com  
*this data was anonymously submitted

The sales dataset contains 1143 observations and has 11 columns with data on:  

 - ad_id: an unique ID for each ad.  
 - xyzcampaignid: an ID associated with each ad campaign of XYZ company.  
 - fbcampaignid: an ID associated with how Facebook tracks each campaign.  
 - age: age of the person to whom the ad is shown.  
 - gender: gender of the person to whim the add is shown  
 - interest: a code specifying the category to which the person’s interest belongs (interests are as mentioned in the person’s Facebook public profile).  
 - Impressions: the number of times the ad was shown.  
 - Clicks: number of clicks on for that ad.  
 - Spent: Amount paid by company xyz to Facebook, to show that ad.  
 - Total conversion: Total number of people who enquired about the product after seeing the ad.  
 - Approved conversion: Total number of people who bought the product after seeing the ad.

```{r}
library(tidyverse)

## importing dataset into r
marketsale <- read_csv("KAG_conversion_data.csv")

## view data
head(marketsale,10)
```
 
 
#### Data Visualization  

First I want to make one static dashboard that breaks down how the company's ads performed on Facebook.

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
## create new data frame for first figure
library(dplyr)

## find the total amount spent for each ad campaign and total number of adds deployed by FB
marketsale %>% group_by(xyz_campaign_id) %>% summarise(sum(Spent)) 
marketsale %>% count(xyz_campaign_id) 

## create data frame
ad.spend = c(150, 2893,55662)
ad.count = c(54,464,625)
ad.number = c("916","936","1178")
ad.info = tibble(ad.number,ad.spend,ad.count)
ad.info
```

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
## create first figure using ad.info data
library(ggplot2)
library(cowplot)
library(scales)

ad.breakdown = ggplot(ad.info) +
  geom_bar(aes(x=reorder(ad.number,ad.count),y=ad.count,fill=reorder(ad.number,ad.count),color=reorder(ad.number,ad.count)),stat = "identity",size=1) +
  geom_line(aes(x=reorder(ad.number,ad.spend),y=ad.spend/75,group=1),size=2,color="#2E2A39",stat = "identity",alpha=0.8) +
  geom_point(aes(x=reorder(ad.number,ad.spend),y=ad.spend/75),color="#2e2a39",size=3) +
  scale_y_continuous(name = "Total Facebook Ads", sec.axis = sec_axis(~.*75,name="Ad Spend ($)",labels = label_number_si())) +
  theme_cowplot(12) + scale_color_manual(values = c("#E3D096","#FD9301","#53929B")) +
  scale_fill_manual(values = alpha(c("#E3D096","#FD9301","#53929B"),0.75)) +
  labs(x="Ad Campaign ID",caption = "The bars represent the total number of Facebook ads deployed\n for each ad campaign. The line and points represent the\n total amount spent on each ad campaign.") +
  theme(legend.position = "none",axis.title = element_text(size=14, face="bold"),axis.text = element_text(size=12), axis.title.y.right = element_text(margin = margin(l=10)), plot.caption = element_text(hjust = 0.5,size=11))
ad.breakdown
```

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
## create graph comparing ad spend to approved conversions
library(scales)

approve = ggplot(marketsale,aes(Spent,Approved_Conversion)) + geom_point(size=4,alpha=0.6) +
  geom_smooth(method = lm,se=FALSE,color="#53929B",size=1.5) + labs(x="Ad Spend", y="Approved Conversions") +
  theme_cowplot(12) + theme(axis.title = element_text(face="bold",size = 12)) + scale_x_continuous(labels = dollar_format())
approve
```

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
## create graph comparing ad spend to total conversions
total.con = ggplot(marketsale,aes(Spent,Total_Conversion)) + geom_point(size=4,alpha=0.6) +
  geom_smooth(method = lm,se=FALSE,color="#53929B",size=1.5) + labs(x="", y="Total Conversions") +
  theme_cowplot(12) + theme(axis.title = element_text(face="bold",size = 12)) +
  scale_x_continuous(labels = dollar_format())
total.con
```

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
## create plots that display the clickthrough rate (CTR) and conversions per impression (CPI) for each ad campaign
ctr.plot = marketsale %>% group_by(xyz_campaign_id) %>% summarise(imp.sum = sum(Impressions),click.sum = sum(Clicks)) %>%
  mutate(CTR = click.sum/imp.sum) %>% ggplot(aes(x=reorder(xyz_campaign_id,CTR),y=CTR)) +
  geom_segment(aes(x=reorder(xyz_campaign_id,CTR),xend=reorder(xyz_campaign_id,CTR), y=0,yend=CTR),color="grey50",size=1) +
  geom_point(size=7, color="#FD9301",fill=alpha("#FD9301",0.5),shape=21,stroke=2) + theme_cowplot(12) +
  labs(x="",y="",title = "Campaigns by Clickthrough Rate (CTR)") +
  theme(plot.title = element_text(size=14,face="bold",hjust = 0.5), axis.text.y = element_text(face="bold")) + coord_flip() +
  scale_y_continuous(labels = percent_format())

cpi.plot = marketsale %>% group_by(xyz_campaign_id) %>% summarise(imp.sum = sum(Impressions),conv.sum = sum(Approved_Conversion)) %>%
  mutate(CPI = conv.sum/imp.sum) %>% ggplot(aes(x=reorder(xyz_campaign_id,CPI),y=CPI)) +
  geom_segment(aes(x=reorder(xyz_campaign_id,CPI),xend=reorder(xyz_campaign_id,CPI), y=0,yend=CPI),color="grey50",size=1) +
  geom_point(size=7, color="#FD9301",fill=alpha("#FD9301",0.5),shape=21,stroke=2) + theme_cowplot(12) + labs(x="",y="",title = "Campaigns by Conversion per Impression (CPI)") +
  theme(plot.title = element_text(size=14,face="bold",hjust = 0.5), axis.text.y = element_text(face="bold")) + coord_flip() + scale_y_continuous(labels = percent_format())

ctr.plot
cpi.plot
```

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
library(ggpubr)

## arrange plots into static dashboard
conv.plots = ggarrange(total.con,approve, ncol = 1,nrow = 2,heights = c(1,1))
top = ggarrange(ad.breakdown,conv.plots,nrow = 1, ncol = 2, widths = c(1,0.6))
bottom = ggarrange(ctr.plot,cpi.plot,nrow = 1,ncol = 2)
web.dash = ggarrange(top, bottom,ncol = 1,nrow = 2, heights = c(1,0.6))
web.final = annotate_figure(web.dash, top = text_grob("Marketing Dashboard - Web Analytics",
            color = "#2e2a39", face = "bold", size = 28, just = "center",lineheight = 2),
            bottom = text_grob("Data source: kaggle.com", color = "#2e2a39", just = "right", x = 1, face = "italic", size = 10))
save_plot("webdash.pdf",web.final,base_height = 10,base_width = 10)
```
You can view the final web analytics dashboard as a pdf on the "PortfolioProjects" repository. It's saved as "webdash.pdf".

---


Next, I want to create a dashboard that looks at customer insights and determines what groups should potentially be targeted in the future.  

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
## create bar chart showing the percentage of approved conversions for each ad based on age group 
customer1 = marketsale %>% group_by(xyz_campaign_id,age) %>% summarise(approve.sum = sum(Approved_Conversion)) %>%
  mutate(percent.approve = approve.sum/sum(approve.sum)) %>%
  ggplot(aes(x= as.factor(xyz_campaign_id), y=percent.approve,fill=age,color=age)) +
  geom_bar(stat = "identity",position = position_dodge()) +
  scale_fill_manual(values=alpha(c("#efd9a7","#c97f04","#617f7f","#192d2e"),0.75),name="Age Group") +
  scale_color_manual(values=c("#efd9a7","#c97f04","#617f7f","#192d2e"),name="Age Group") +
  theme_cowplot(12) + scale_y_continuous(labels = scales::percent) +
  labs(x="Ad Campaign ID",y="Percent of Approved Conversions",title = "Percent of Approved Conversions by\n Campaign and Age Group") +
  theme(plot.title = element_text(size=14,face="bold",hjust = 0.5), legend.title = element_text(size=12, face="bold"), legend.text = element_text(size=10), axis.title = element_text(face="bold"))

customer1
```
```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
## change M and F in gender column to Male and Female
marketsale$gender[marketsale$gender == "F"] <- "Female"
marketsale$gender[marketsale$gender == "M"] <- "Male"

## create graph breaking down percent of approved conversions by gender
customer2 = marketsale %>% group_by(gender) %>% summarise(approve.sum = sum(Approved_Conversion)) %>%
  mutate(g.convert = approve.sum/sum(approve.sum)*100) %>% ggplot(aes(x="", y=g.convert, fill=gender,color=gender)) +
  geom_bar(stat="identity", width=1,size=1) + coord_polar("y", start=0) + theme_void() +
  theme(legend.position="none") + scale_fill_manual(values = alpha(c("#c97f04","#192d2e"),0.8)) +
  scale_color_manual(values = c("#c97f04","#192d2e")) +
  geom_text(aes(label = paste0(round(g.convert),"% ", gender)), color = "black", size=4, show.legend = FALSE,position = position_stack(vjust = 0.5)) +
  labs(title = "Approved Conversions by Gender\n for All Ad Campaigns") +
  theme(plot.title = element_text(size=14, face="bold", hjust = 0.5))

customer2
```

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
## create graph showing total number of clicks based on customer interests (on Facebook)
clicks.int = ggplot(marketsale, aes(x=interest,y=Clicks)) + geom_point(color="#617f7f",alpha=0.7,size=3) + theme_cowplot(12) +
  labs(x="Customer Interests",title = "Number of Clicks based on\n Customer Interests") +
  theme(axis.title = element_text(size=12, face="bold"),plot.title = element_text(hjust = 0.5))

## create graph showing munber of approved interactions based on customer interests (on Facebook)
approve.int = ggplot(marketsale, aes(x=interest,y=Approved_Conversion)) + 
  geom_point(color="#c97f04",alpha=0.7,size=3) + theme_cowplot(12) + 
  labs(x="Customer Interests",y="Approved Conversions",title = "Number of Approved Conversions\n based on Customer Interests") +
  theme(axis.title = element_text(size=12, face="bold"),plot.title = element_text(hjust = 0.5))

clicks.int
approve.int
```
```{r}
## arrange plots into static dashboard
customer.top = ggarrange(customer1,customer2,nrow = 1,ncol = 2,widths = c(1,0.5))
customer.bottom = ggarrange(clicks.int,approve.int,nrow = 1,ncol = 2)
customer.dash = ggarrange(customer.top, customer.bottom,ncol = 1,nrow = 2, heights = c(1,0.7))
customer.final = annotate_figure(customer.dash, top = text_grob("Marketing Dashboard - Customer Insights", color = "#2e2a39", face = "bold", size = 28, just = "center",lineheight = 2), bottom = text_grob("Data source: kaggle.com", color = "#2e2a39", just = "right", x = 1, face = "italic", size = 10))
save_plot("customerdash.pdf",customer.final,base_height = 10,base_width = 10)
```
You can view the final customer insights  dashboard as a pdf on the "PortfolioProjects" repository. It's saved as "customerdash.pdf".

